### 2018-06-12 paper/5-implementations.tex
Work on Make and Excel implementations

% Type signature for process. Not sure we really want this, as it's quite long
% and is repeated in topological.
% process :: k -> [k] -> State@\,@(Store@\,@(MakeInfo@\,\blk{k}@)@\,\blk{k}@@\,\blk{v}@)@\,\blk{v}@ -> State@\,@(Store@\,@(MakeInfo@\,\blk{k}@)@\,\blk{k}@@\,\blk{v}@)@\,@(@@)
\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Persistent build information
type Time       = Integer
type MakeInfo k = (k -> Time, Time)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Make build system
make :: Eq k => Build Applicative (MakeInfo k) k v
make = topological process
  where
    process key deps act = do
        (modTime, now) <- gets getInfo
        let dirty = or [ modTime dep > modTime key | dep <- deps ]
        when dirty $ do
            v <- act
            let newModTime k = if k == key then now else modTime k
            modify $ putInfo (newModTime, now + 1) . putValue key v
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard graph algorithms (implementation omitted)
reachable :: Eq a => (a -> [a]) ->  a  -> [a]
topSort   :: Eq a => (a -> [a]) -> [a] -> [a]
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Topological dependency strategy
topological :: Eq k => (k -> [k] -> State (Store i k v) v -> State (Store i k v) ())
            -> Build Applicative i k v
topological process task key = execState $ forM_ chain $ \k -> do
    let fetch k = gets (getValue k)
    case task fetch k of
        Nothing  -> return ()
        Just act -> process k (deps k) act
  where
    deps  = dependencies task -- dependencies is defined in @\S\ref{secdeps}@
    chain = topSort deps (reachable deps key)
\end{minted}
\vspace{-2mm}
\caption{An implementation of \Make using our framework.}\label{fig-make-implementation}
\vspace{-2mm}
\end{figure}

@@

We define \Excel, with its reordering dependency strategy, in  Fig.~\ref{fig-excel-implementation}.
\Excel's persistently stored information is a triple: (i) the dirty bit
function \hs{k}~\hs{->}~\hs{Bool}, (ii) an approximation of key dependencies
\hs{k}~\hs{->}~\hs{DependencyApproximation}~\hs{k} that \Excel uses to handle
dynamic dependencies, and (iii) the calculation chain \hs{[@@k]} recorded in the
previous build (\S\ref{sec-background-excel}).

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Approximation of task dependencies
data DependencyApproximation k = SubsetOf [k] | Unknown
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Persistent build information
type CalcChain k = [k]
type ExcelInfo k = ((k -> Bool, k -> DependencyApproximation k), CalcChain k)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Result of speculative task execution
data Result k v = MissingDependency k | Result v [k]
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Reordering dependency strategy (implementation omitted, 21 lines)
reordering :: Ord k
    => (k -> State (Store i k v) (Result k v) -> State (Store i k v) (Maybe (Result k v)))
    -> Build Monad (i, CalcChain k) k v
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Excel build system
excel :: Ord k => Build Monad (ExcelInfo k) k v
excel = reordering process
  where
    process key act = do
        (dirty, deps) <- gets getInfo
        let rebuild = dirty key || case deps key of SubsetOf ks -> any dirty ks
                                                    Unknown     -> True
        if not rebuild
            then return Nothing
            else do
                result <- act
                case result of
                    MissingDependency _ -> return ()
                    Result v dynamicDependencies -> do
                        let newDirty k = if k == key then True else dirty k
                        modify $ putInfo (newDirty, deps) . putValue key v
                return (Just result)
\end{minted}
\vspace{-2mm}
\caption{An implementation of \Excel using our framework.}\label{fig-excel-implementation}
\vspace{-2mm}
\end{figure}

>>>>

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Build information: current time and file modification times
type Time       = Integer
type MakeInfo k = (Time, Map k Time)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Make build system
make :: Ord k => Build Applicative (MakeInfo k) k v
make = topological modTimeRebuilder
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Topological scheduler
topological :: Ord k => Rebuilder Applicative i k v -> Build Applicative i k v
topological rebuilder tasks key = execState $ forM_ chain $ \k -> case tasks k of
    Nothing   -> return ()
    Just task -> do
        value <- gets (getValue k)
        let newTask :: Task (MonadState i) k v
            newTask = rebuilder k value task
            newFetch :: k -> StateT i (State (Store i k v)) v
            newFetch = lift . gets . getValue
        info <- gets getInfo
        (newValue, newInfo) <- runStateT (run newTask newFetch) info
        modify $ putInfo newInfo . putValue k newValue
  where
    deps k = case tasks k of { Nothing -> []; Just task -> dependencies task }
    chain  = case topSort (reachable deps key) of
        Nothing -> error "Cannot build tasks with cyclic dependencies"
        Just xs -> xs
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard graph algorithms (implementation omitted)
reachable :: Ord k => (k -> [k]) -> k -> Graph k
topSort   :: Ord k => Graph k -> Maybe [k]
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on file modification times
modTimeRebuilder :: Ord k => Rebuilder Applicative (MakeInfo k) k v
modTimeRebuilder key value task = Task $ \fetch -> do
    (now, modTime) <- get
    let dirty = case Map.lookup key modTime of
            Nothing -> True
            time -> any (\d -> Map.lookup d modTime > time) (dependencies task)
    if not dirty
    then return value
    else do
        put (now + 1, Map.insert key now modTime)
        run task fetch
\end{minted}
\vspace{-2mm}
\caption{An implementation of \Make using our framework.}\label{fig-make-implementation}
\vspace{-2mm}
\end{figure}

@@

We define \Excel, with its reordering dependency strategy, in  Fig.~\ref{fig-excel-implementation}.
\Excel's persistently stored information is a triple: (i) the dirty bit
function \hs{k}~\hs{->}~\hs{Bool}, (ii) an approximation of key dependencies
\hs{k}~\hs{->}~\hs{ApproximateDependencies}~\hs{k} that \Excel uses to handle
dynamic dependencies, and (iii) the calculation chain \hs{[@@k]} recorded in the
previous build (\S\ref{sec-background-excel}).

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Approximation of task dependencies: a map of known upper bounds of dependencies
type ApproximateDependencies k = Map k [k]
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build information: dirty keys, dependency approximation, and calculation chain
type CalcChain k = [k]
type ExcelInfo k = ((Set k, ApproximateDependencies k), CalcChain k)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Excel build system
excel :: Ord k => Build Monad (ExcelInfo k) k v
excel = restarting approximateRebuilder
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Restarting scheduler
restarting :: Ord k => Rebuilder Monad i k v -> Build Monad (i, Chain k) k v
restarting rebuilder tasks key = execState $ do
    chain    <- gets (snd . getInfo)
    newChain <- go Set.empty $ chain ++ [key | key `notElem` chain]
    modify . mapInfo $ \(i, _) -> (i, newChain)
  where
    go :: Set k -> Chain k -> State (Store (i, Chain k) k v) (Chain k)
    go _    []     = return []
    go done (k:ks) = case tasks k of
        Nothing -> (k :) <$> go (Set.insert k done) ks
        Just task -> do
            store <- get
            let value = getValue k store
                newTask :: Task (MonadState i) k (Either k v)
                newTask = try $ rebuilder k value task -- `try' is defined below
                newFetch :: k -> State i (Either k v)
                newFetch k | k `Set.member` done = return $ Right (getValue k store)
                           | otherwise           = return $ Left k
            case runState (run newTask newFetch) (fst $ getInfo store) of
                (Left dep, _) -> go done $ [ dep | dep `notElem` ks ] ++ ks ++ [k]
                (Right newValue, newInfo) -> do
                    modify $ putInfo (newInfo, []) . updateValue k value newValue
                    (k :) <$> go (Set.insert k done) ks
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Convert a total task into a task accepting a partial fetch callback
try :: Task (MonadState i) k v -> Task (MonadState i) k (Either e v)
try task = Task $ \fetch -> runExceptT $ run task (ExceptT . fetch)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Rebuilder based on approximate dependencies
approximateRebuilder :: Ord k => Rebuilder Monad (Set k, ApproximateDependencies k) k v
approximateRebuilder key value task = Task $ \fetch -> do
    (isDirty, deps) <- get
    let dirty = key `Set.member` isDirty || case Map.lookup key deps of
                                                Nothing -> True
                                                Just ks -> any (`Set.member` isDirty) ks
    if not dirty
    then return value
    else do
        put (Set.insert key isDirty, deps)
        run task fetch
\end{minted}
\vspace{-2mm}
\caption{An implementation of \Excel using our framework.}\label{fig-excel-implementation}
\vspace{-2mm}
\end{figure}
